New and simplified manual controls for projection and slice tours, with application to exploring classification boundaries in high dimensions

### 2023-02-08 paper/appendix.Rmd
Last few changes

\textcolor{blue}{The slice center position is controlled via numerical input along each axis in the original command `SliceDynamic`. This provides numerical feedback on the position of the center. For easier navigation but without the numerical feedback, an implementation of the slice center control via a visual display like the one in Figure \ref{fig:anchornav} is provided in `SliceDynamicCC`.}

>>>>

The slice center position is controlled via numerical input along each axis in the original command `SliceDynamic`. This provides numerical feedback on the position of the center. For easier navigation but without the numerical feedback, an implementation of the slice center control via a visual display like the one in Figure \ref{fig:anchornav} is provided in `SliceDynamicCC`.

### 2023-02-08 paper/appendix.tex
Last few changes

\textcolor{blue}{The slice center position is controlled via numerical input along each axis in the original command `SliceDynamic`. This provides numerical feedback on the position of the center. For easier navigation but without the numerical feedback, an implementation of the slice center control via a visual display like the one in Figure \ref{fig:anchornav} is provided in `SliceDynamicCC`.}

>>>>

The slice center position is controlled via numerical input along each
axis in the original command \texttt{SliceDynamic}. This provides
numerical feedback on the position of the center. For easier navigation
but without the numerical feedback, an implementation of the slice
center control via a visual display like the one in Figure
\ref{fig:anchornav} is provided in \texttt{SliceDynamicCC}.

### 2023-02-08 paper/paper.Rmd
Last few changes

<!-- Background -->
From a statistical perspective 3D is a rare data dimension, so unlike in most 3D rotation computer graphics applications, \textcolor{blue}{methods which work for arbitrary dimensions have more general utility}. A good approach is to show projections from an arbitrary dimensional space to create dynamic data visualizations called *tours*. Tours involve views of high-dimensional data with low-dimensional projections. In his original paper on the grand tour, @As85 provided several algorithms for tour paths that could theoretically show the viewer the data *from all sides*. \textcolor{blue}{(Various videos of tours, e.g.} @dataviewer, @gt-pp-video, on the @ASA22, \textcolor{blue}{illustrate this concept.)} Prior to Asimov's work, there were numerous preparatory developments including @tukey's PRIM-9. PRIM-9 had user-controlled rotations on coordinate axes, allowing one to manually tour through low-dimensional projections. (A video [@PRIM9-video] illustrating the capabilities is available through \textcolor{blue}{the} video library of @ASA22.) Steering through all possible projections is impossible, unlike Asimov's tours which allow one to quickly see many, many different projections. After Asimov, there have been many tour developments, which are summarized in @lee2021.

@@

A manual tour allows the user to alter the coefficients of one (or more) variables, of $p$, contributing to a $d$-dimensional projection. The initial ingredients are an orthonormal basis ($A_{p\times d}$) defining the projection of the data, and a variable id ($m \in \{1, ..., p\}$) specifying which coefficient will be \textcolor{blue}{changed: the $m^{th}$ row of $A_{p\times d}$ which we denote by $V_m$.} A method to update the values of the component of the controlled variable $V_m$ is then needed.

@@

1. Provide $A$, and $m$. (Note that $m$ could also be automatically chosen as the component that is closest to the cursor position.)
2. Change values in row $m$, for example, if $d=2$ 
$$
A^* = [ \boldsymbol{a}^*_1~\boldsymbol{a}^*_2 ] = \left[ \begin{array}{cc} a_{11} & a_{12}\\
                              \vdots & \vdots \\
                              a^*_{m1} & a^*_{m2}\\
                              \vdots & \vdots \\
                              a_{p1} & a_{p2} 
        \end{array}\right].
$$ 
\noindent A large change in these values would correspond to making a large jump from the current projection. Small changes would correspond to tracking a cursor, making small jumps from the current projection.
3. Orthonormalise $A^*$, using Gram-Schmidt. 
    i. Normalise $\boldsymbol{a}^*_1$ and $\boldsymbol{a}^*_2$.
    ii. $\boldsymbol{a}^*_2 = \boldsymbol{a}^*_2 - (\boldsymbol{a}^*_1\cdot\boldsymbol{a}^*_2)\boldsymbol{a}^*_1$.

\textcolor{red}{a2 should be normalised after the second step? Fig 1: this is now showing tracking, with the yellow dot moving in each image?}

@@

```{r manualsequence, out.width="100%", width=12, height=4, fig.cap="Sequence of projections where the contribution of one variable is controlled (dark) is changed using unconstrained orthonormalization. The dot indicates the chosen values for the controlled variable, $V_m$. It can be seen that the actual axis does not precisely match the chosen position, but it is close."}
source("../code/plot_basis.R")
source("../code/linear_alg.R")

@@

The implementation of the manual tour as suggested here requires the visualization of the current projection in terms of an axis display (see Figure \ref{fig:manualsequence} for an example). This display should be able to track the mouse position and adjust the projection based on the \textcolor{blue}{user's cursor position}. The user can select one axis (corresponding to one variable) by clicking on it, and then adjust the position of that axis by dragging it. In practice, the closest axis will be selected, and the dragging results in tracking the mouse position which will update the projection in small steps, such that no further interpolation is required.

The interactions in the axis display need to be mapped back onto the current projection matrix, which will then be orthonormalized and \textcolor{blue}{re-displayed}. A second, linked display shows the projected data in sync with the updates from the axis display.

@@

\textcolor{blue}{Within R, the natural implementation for an interactive interface would use} `shiny` [@shinym] with `plotly` [@plotly] drawing. However, the `plotly` package is designed to add to `ggplot2` [@ggplot2] graphics, not to be a programmable interactive graphics system. It lacks the capability to implement the primary tool needed for a manual tour, which is the tracking of the cursor position in small increments. In theory, the historical `locator()` function of base R should provide the necessary capability, but its behavior is not consistent between graphical devices. The cursor position control is possible for 3D plots with `rgl` [@rgl] and `plotly` but these not the arbitrary dimension needed for tours. The level of interactivity would have been possible with the `cranvas` system [@cranvas] or `rJava` [@rJava] but neither has a graphics toolkit that currently operates across platforms.  What we have chosen to do is to use the interactivity of Mathematica to experiment, and to implement the resulting ideas as an animation in the `tourr` [@tourr] package. There is hope that better interactivity will be available for R soon, perhaps in javascript [e.g. @langevitour, @detourr], which might allow for the experimentation to develop new techniques.

@@

\textcolor{blue}{The Mathematica implementation relies on the inbuilt function  `LocatorPane`. This function creates a region on the screen where the position of the cursor is captured and then converted to input that updates the graphics functions, enabling the manual navigation of the tour. We then use the graphics function `ListPlot` embedded in dynamic objects via  `DynamicModule` to produce the display. User interface objects, such as sliders, locator panes, and input fields, are used to provide user control of slice thickness, center, scale, and point size.}

The examples presented in the application section below (and attached in notebook format in the supplementary material) all use our primary new function, `SliceDynamic`. This function typically accepts grouped data in the form of a matrix where the second last column details the name of the group and the last column details the group index (which can be one if there is only one group). After specifying the initial slice thickness and the slice range, the user is presented with an interactive display in which the control objects appear on the left and the slice visualization appears on the right. The user can change the orientation of the slice via the locator pane, which changes the projection matrix;  a slider controls the slice thickness; and there is an input field, which changes the slice center.
\textcolor{red}{insert sentence about new slice position display here.}
The user can also change the appearance of the plot by zooming into the center or changing the point size with the sliders provided. It is worth noting that this zoom works best with data that has been centered and scaled. The projected data can be displayed to contrast it with the slice by ticking a box. The explicit projection matrix can also be displayed via another checkbox. This is especially useful when we wish to import a projection that was identified as interesting in the manual exploration into a later study, as shown in the applications below. More details about the implementation and usage instructions are given in the Appendix.

@@

We start with a comparison of the two models and the data distribution shifting to $\mathbf{c}^{+}$, thus the slice is localized towards high values of bill depth in the top row of Figure \ref{slice1p}. \textcolor{blue}{The star glyph at the right indicates the slice center value, which shows that it is higher on `bd` and at the middle value for the other variables. This slice center guide is as proposed in @condviz2, and more details can be found in the Appendix.} We can see that all three slices (the two models and the data) contain almost no points from the third class (Gentoo) and that the decision boundary between the two models is very similar.

@@

A more interesting comparison is found for $\mathbf{c}^{-}$, thus the slice localized towards low values of bill depth, shown in the bottom row of Figure \ref{slice1p}. \textcolor{blue}{(The star glyph at right indicates the slice center value, which shows that it is lower on `bd` and at the middle value for the other variables.)} The RF model (left) predicts all three species within this slice, with an interesting boundary for the third class (Gentoo). On the other hand, the LDA model (middle) predominantly predicts the third class within the slice, this appears to be enforced through the linear structure of the model. Looking finally at the thick slice through the data we see that there are primarily observations from this class within the slice we can conclude that the two models have filled in the "empty" space (where we do not have any training observations) in very different ways and according to what we might expect given the model structure.

@@

Mathematica provided a useful sandbox to experiment with the ideas presented in this paper. 
Most of the data, though, is first constructed using R, especially the classification boundaries. \textcolor{red}{what should this say?}
If new technology for interactive graphics with precise cursor control becomes available in R it would be useful to create a tighter coupling of models and visualization to allow exploring and comparing fits.

@@

- The Mathematica source code defining the new functions in mmtour.wl.
- Two Mathematica notebooks with the code used in the application (corresponding to the three subsections).
- Appendix with additional details about manual controls, and Mathematica functions.
- Data files for the examples.
- R code to reproduce the example data, Figures 1 and 2 in the paper and Figures 1 and 2 in the appendix.
- Animations illustrating the manual tour and slicing are also available at https://vimeo.com/747585410 and https://vimeo.com/747590472.

>>>>

<!-- Background -->
From a statistical perspective 3D is a rare data dimension, so unlike in most 3D rotation computer graphics applications, methods which work for arbitrary dimensions have more general utility. A good approach is to show projections from an arbitrary dimensional space to create dynamic data visualizations called *tours*. Tours involve views of high-dimensional data with low-dimensional projections. In his original paper on the grand tour, @As85 provided several algorithms for tour paths that could theoretically show the viewer the data *from all sides*. (Various videos of tours, e.g. @dataviewer, @gt-pp-video, on the @ASA22, illustrate this concept.) Prior to Asimov's work, there were numerous preparatory developments including @tukey's PRIM-9. PRIM-9 had user-controlled rotations on coordinate axes, allowing one to manually tour through low-dimensional projections. (A video [@PRIM9-video] illustrating the capabilities is available through the video library of @ASA22.) Steering through all possible projections is impossible, unlike Asimov's tours which allow one to quickly see many, many different projections. After Asimov, there have been many tour developments, which are summarized in @lee2021.

@@

A manual tour allows the user to alter the coefficients of one (or more) variables, of $p$, contributing to a $d$-dimensional projection. The initial ingredients are an orthonormal basis ($A_{p\times d}$) defining the projection of the data, and a variable id ($m \in \{1, ..., p\}$) specifying which coefficient will be changed: the $m^{th}$ row of $A_{p\times d}$ which we denote by $V_m$. A method to update the values of the component of the controlled variable $V_m$ is then needed.

@@

1. Provide $A$, and $m$. (Note that $m$ could also be automatically chosen as the component that is closest to the cursor position.)
2. Change values in row $m$, for example, if $d=2$ 
$$
A^* = [ \boldsymbol{a}^*_1~\boldsymbol{a}^*_2 ] = \left[ \begin{array}{cc} a_{11} & a_{12}\\
                              \vdots & \vdots \\
                              a^*_{m1} & a^*_{m2}\\
                              \vdots & \vdots \\
                              a_{p1} & a_{p2} 
        \end{array}\right].
$$ 
\noindent A large change in these values would correspond to making a large jump from the current projection. Small changes would correspond to tracking a cursor, making small jumps from the current projection.
3. Orthonormalise $A^*$, using Gram-Schmidt. 
    i. Normalise $\boldsymbol{a}^*_1$.
    ii. $\boldsymbol{a}^*_2 = \boldsymbol{a}^*_2 - (\boldsymbol{a}^*_1\cdot\boldsymbol{a}^*_2)\boldsymbol{a}^*_1$.
    iii. Normalise $\boldsymbol{a}^*_2$.

@@

```{r manualsequence, out.width="100%", width=12, height=4, fig.cap="Sequence of projections where the contribution of one variable is controlled (dark). The dot indicates the chosen values for the controlled variable, $V_m$. The change is tracking the chosen value after unconstrained orthonormalization. It can be seen that the actual axis does not precisely match the chosen position, but it is close."}
source("../code/plot_basis.R")
source("../code/linear_alg.R")

@@

The implementation of the manual tour as suggested here requires the visualization of the current projection in terms of an axis display (see Figure \ref{fig:manualsequence} for an example). This display should be able to track the mouse position and adjust the projection based on the user's cursor position. The user can select one axis (corresponding to one variable) by clicking on it, and then adjust the position of that axis by dragging it. In practice, the closest axis will be selected, and the dragging results in tracking the mouse position which will update the projection in small steps, such that no further interpolation is required.

The interactions in the axis display need to be mapped back onto the current projection matrix, which will then be orthonormalized and re-displayed. A second, linked display shows the projected data in sync with the updates from the axis display.

@@

Within R, the natural implementation for an interactive interface would use `shiny` [@shinym] with `plotly` [@plotly] drawing. However, the `plotly` package is designed to add to `ggplot2` [@ggplot2] graphics, not to be a programmable interactive graphics system. It lacks the capability to implement the primary tool needed for a manual tour, which is the tracking of the cursor position in small increments. In theory, the historical `locator()` function of base R should provide the necessary capability, but its behavior is not consistent between graphical devices. The cursor position control is possible for 3D plots with `rgl` [@rgl] and `plotly` but these not the arbitrary dimension needed for tours. The level of interactivity would have been possible with the `cranvas` system [@cranvas] or `rJava` [@rJava] but neither has a graphics toolkit that currently operates across platforms.  What we have chosen to do is to use the interactivity of Mathematica to experiment, and to implement the resulting ideas as an animation in the `tourr` [@tourr] package. There is hope that better interactivity will be available for R soon, perhaps in javascript [e.g. @langevitour, @detourr], which might allow for the experimentation to develop new techniques.

@@

The Mathematica implementation relies on the inbuilt function  `LocatorPane`. This function creates a region on the screen where the position of the cursor is captured and then converted to input that updates the graphics functions, enabling the manual navigation of the tour. We then use the graphics function `ListPlot` embedded in dynamic objects via  `DynamicModule` to produce the display. User interface objects, such as sliders, locator panes, and input fields, are used to provide user control of slice thickness, center, scale, and point size.

The examples presented in the application section below (and attached in notebook format in the supplementary material) all use our primary new function, `SliceDynamic`. This function typically accepts grouped data in the form of a matrix where the second last column details the name of the group and the last column details the group index (which can be one if there is only one group). After specifying the initial slice thickness and the slice range, the user is presented with an interactive display in which the control objects appear on the left and the slice visualization appears on the right. The user can change the orientation of the slice via the locator pane, which changes the projection matrix;  a slider controls the slice thickness; and there is an input field or an interactive star glyph, to change the slice center. The user can also change the appearance of the plot by zooming into the center or changing the point size with the sliders provided. It is worth noting that this zoom works best with data that has been centered and scaled. The projected data can be displayed to contrast it with the slice by ticking a box. The explicit projection matrix can also be displayed via another checkbox. This is especially useful when we wish to import a projection that was identified as interesting in the manual exploration into a later study, as shown in the applications below. More details about the implementation and usage instructions are given in the Appendix.

@@

We start with a comparison of the two models and the data distribution shifting to $\mathbf{c}^{+}$, thus the slice is localized towards high values of bill depth in the top row of Figure \ref{slice1p}. The star glyph at the right indicates the slice center value, which shows that it is higher on `bd` and at the middle value for the other variables. This slice center guide is as proposed in @condviz2, and more details can be found in the Appendix. We can see that all three slices (the two models and the data) contain almost no points from the third class (Gentoo) and that the decision boundary between the two models is very similar.

@@

A more interesting comparison is found for $\mathbf{c}^{-}$, thus the slice localized towards low values of bill depth, shown in the bottom row of Figure \ref{slice1p}. (The star glyph at right indicates the slice center value, which shows that it is lower on `bd` and at the middle value for the other variables.) The RF model (left) predicts all three species within this slice, with an interesting boundary for the third class (Gentoo). On the other hand, the LDA model (middle) predominantly predicts the third class within the slice, this appears to be enforced through the linear structure of the model. Looking finally at the thick slice through the data we see that there are primarily observations from this class within the slice we can conclude that the two models have filled in the "empty" space (where we do not have any training observations) in very different ways and according to what we might expect given the model structure.

@@

Mathematica provided a useful sandbox to experiment with the ideas presented in this paper. However, we have used R for the statistical data analysis and modeling in our example, as is the standard approach in the community. Since the proposed visualization should ideally be integrated within this workflow, the new methods should also be implemented as interactive graphics accessible from R once new technology with precise cursor control becomes available.

@@

- The Mathematica source code defining the new functions in mmtour.wl.
- Two Mathematica notebooks with the code used in the application.
- Appendix with additional details about manual controls, and Mathematica functions.
- Data files for the examples.
- R code to reproduce the example data, Figures 1 and 2 in the paper and Figures 1 and 2 in the appendix.
- Animations illustrating the manual tour and slicing are also available at https://vimeo.com/747585410 and https://vimeo.com/747590472.

### 2023-02-08 paper/paper.tex
Last few changes

From a statistical perspective 3D is a rare data dimension, so unlike in
most 3D rotation computer graphics applications,
\textcolor{blue}{methods which work for arbitrary dimensions have more general utility}.
A good approach is to show projections from an arbitrary dimensional
space to create dynamic data visualizations called \emph{tours}. Tours
involve views of high-dimensional data with low-dimensional projections.
In his original paper on the grand tour, \citet{As85} provided several
algorithms for tour paths that could theoretically show the viewer the
data \emph{from all sides}.
\textcolor{blue}{(Various videos of tours, e.g.} \citet{dataviewer},
\citet{gt-pp-video}, on the \citet{ASA22},
\textcolor{blue}{illustrate this concept.)} Prior to Asimov's work,
there were numerous preparatory developments including \citet{tukey}'s
PRIM-9. PRIM-9 had user-controlled rotations on coordinate axes,
allowing one to manually tour through low-dimensional projections. (A
video \citep{PRIM9-video} illustrating the capabilities is available
through \textcolor{blue}{the} video library of \citet{ASA22}.) Steering
through all possible projections is impossible, unlike Asimov's tours
which allow one to quickly see many, many different projections. After
Asimov, there have been many tour developments, which are summarized in
\citet{lee2021}.

@@

A manual tour allows the user to alter the coefficients of one (or more)
variables, of \(p\), contributing to a \(d\)-dimensional projection. The
initial ingredients are an orthonormal basis (\(A_{p\times d}\))
defining the projection of the data, and a variable id
(\(m \in \{1, ..., p\}\)) specifying which coefficient will be
\textcolor{blue}{changed: the $m^{th}$ row of $A_{p\times d}$ which we denote by $V_m$.}
A method to update the values of the component of the controlled
variable \(V_m\) is then needed.

@@

  \begin{enumerate}
  \def\labelenumii{\roman{enumii}.}
  \tightlist
  \item
    Normalise \(\boldsymbol{a}^*_1\) and \(\boldsymbol{a}^*_2\).
  \item
    \(\boldsymbol{a}^*_2 = \boldsymbol{a}^*_2 - (\boldsymbol{a}^*_1\cdot\boldsymbol{a}^*_2)\boldsymbol{a}^*_1\).
  \end{enumerate}
\end{enumerate}

\textcolor{red}{a2 should be normalised after the second step? Fig 1: this is now showing tracking, with the yellow dot moving in each image?}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{paper_files/figure-latex/manualsequence-1} \caption{Sequence of projections where the contribution of one variable is controlled (dark) is changed using unconstrained orthonormalization. The dot indicates the chosen values for the controlled variable, $V_m$. It can be seen that the actual axis does not precisely match the chosen position, but it is close.}\label{fig:manualsequence}
\end{figure}

@@

\textcolor{blue}{Within R, the natural implementation for an interactive interface would use}
\texttt{shiny} \citep{shinym} with \texttt{plotly} \citep{plotly}
drawing. However, the \texttt{plotly} package is designed to add to
\texttt{ggplot2} \citep{ggplot2} graphics, not to be a programmable
interactive graphics system. It lacks the capability to implement the
primary tool needed for a manual tour, which is the tracking of the
cursor position in small increments. In theory, the historical
\texttt{locator()} function of base R should provide the necessary
capability, but its behavior is not consistent between graphical
devices. The cursor position control is possible for 3D plots with
\texttt{rgl} \citep{rgl} and \texttt{plotly} but these not the arbitrary
dimension needed for tours. The level of interactivity would have been
possible with the \texttt{cranvas} system \citep{cranvas} or
\texttt{rJava} \citep{rJava} but neither has a graphics toolkit that
currently operates across platforms. What we have chosen to do is to use
the interactivity of Mathematica to experiment, and to implement the
resulting ideas as an animation in the \texttt{tourr} \citep{tourr}
package. There is hope that better interactivity will be available for R
soon, perhaps in javascript \citep[e.g.][\citet{detourr}]{langevitour},
which might allow for the experimentation to develop new techniques.

@@

\textcolor{blue}{The Mathematica implementation relies on the inbuilt function  `LocatorPane`. This function creates a region on the screen where the position of the cursor is captured and then converted to input that updates the graphics functions, enabling the manual navigation of the tour. We then use the graphics function `ListPlot` embedded in dynamic objects via  `DynamicModule` to produce the display. User interface objects, such as sliders, locator panes, and input fields, are used to provide user control of slice thickness, center, scale, and point size.}

The examples presented in the application section below (and attached in
notebook format in the supplementary material) all use our primary new
function, \texttt{SliceDynamic}. This function typically accepts grouped
data in the form of a matrix where the second last column details the
name of the group and the last column details the group index (which can
be one if there is only one group). After specifying the initial slice
thickness and the slice range, the user is presented with an interactive
display in which the control objects appear on the left and the slice
visualization appears on the right. The user can change the orientation
of the slice via the locator pane, which changes the projection matrix;
a slider controls the slice thickness; and there is an input field,
which changes the slice center.
\textcolor{red}{insert sentence about new slice position display here.}
The user can also change the appearance of the plot by zooming into the
center or changing the point size with the sliders provided. It is worth
noting that this zoom works best with data that has been centered and
scaled. The projected data can be displayed to contrast it with the
slice by ticking a box. The explicit projection matrix can also be
displayed via another checkbox. This is especially useful when we wish
to import a projection that was identified as interesting in the manual
exploration into a later study, as shown in the applications below. More
details about the implementation and usage instructions are given in the
Appendix.

@@

We start with a comparison of the two models and the data distribution
shifting to \(\mathbf{c}^{+}\), thus the slice is localized towards high
values of bill depth in the top row of Figure \ref{slice1p}.
\textcolor{blue}{The star glyph at the right indicates the slice center value, which shows that it is higher on `bd` and at the middle value for the other variables. This slice center guide is as proposed in @condviz2, and more details can be found in the Appendix.}
We can see that all three slices (the two models and the data) contain
almost no points from the third class (Gentoo) and that the decision
boundary between the two models is very similar.

@@

A more interesting comparison is found for \(\mathbf{c}^{-}\), thus the
slice localized towards low values of bill depth, shown in the bottom
row of Figure \ref{slice1p}.
\textcolor{blue}{(The star glyph at right indicates the slice center value, which shows that it is lower on `bd` and at the middle value for the other variables.)}
The RF model (left) predicts all three species within this slice, with
an interesting boundary for the third class (Gentoo). On the other hand,
the LDA model (middle) predominantly predicts the third class within the
slice, this appears to be enforced through the linear structure of the
model. Looking finally at the thick slice through the data we see that
there are primarily observations from this class within the slice we can
conclude that the two models have filled in the ``empty'' space (where
we do not have any training observations) in very different ways and
according to what we might expect given the model structure.

>>>>

From a statistical perspective 3D is a rare data dimension, so unlike in
most 3D rotation computer graphics applications, methods which work for
arbitrary dimensions have more general utility. A good approach is to
show projections from an arbitrary dimensional space to create dynamic
data visualizations called \emph{tours}. Tours involve views of
high-dimensional data with low-dimensional projections. In his original
paper on the grand tour, \citet{As85} provided several algorithms for
tour paths that could theoretically show the viewer the data \emph{from
all sides}. (Various videos of tours, e.g. \citet{dataviewer},
\citet{gt-pp-video}, on the \citet{ASA22}, illustrate this concept.)
Prior to Asimov's work, there were numerous preparatory developments
including \citet{tukey}'s PRIM-9. PRIM-9 had user-controlled rotations
on coordinate axes, allowing one to manually tour through
low-dimensional projections. (A video \citep{PRIM9-video} illustrating
the capabilities is available through the video library of
\citet{ASA22}.) Steering through all possible projections is impossible,
unlike Asimov's tours which allow one to quickly see many, many
different projections. After Asimov, there have been many tour
developments, which are summarized in \citet{lee2021}.

@@

A manual tour allows the user to alter the coefficients of one (or more)
variables, of \(p\), contributing to a \(d\)-dimensional projection. The
initial ingredients are an orthonormal basis (\(A_{p\times d}\))
defining the projection of the data, and a variable id
(\(m \in \{1, ..., p\}\)) specifying which coefficient will be changed:
the \(m^{th}\) row of \(A_{p\times d}\) which we denote by \(V_m\). A
method to update the values of the component of the controlled variable
\(V_m\) is then needed.

@@

  \begin{enumerate}
  \def\labelenumii{\roman{enumii}.}
  \tightlist
  \item
    Normalise \(\boldsymbol{a}^*_1\).
  \item
    \(\boldsymbol{a}^*_2 = \boldsymbol{a}^*_2 - (\boldsymbol{a}^*_1\cdot\boldsymbol{a}^*_2)\boldsymbol{a}^*_1\).
  \item
    Normalise \(\boldsymbol{a}^*_2\).
  \end{enumerate}
\end{enumerate}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{paper_files/figure-latex/manualsequence-1} \caption{Sequence of projections where the contribution of one variable is controlled (dark). The dot indicates the chosen values for the controlled variable, $V_m$. The change is tracking the chosen value after unconstrained orthonormalization. It can be seen that the actual axis does not precisely match the chosen position, but it is close.}\label{fig:manualsequence}
\end{figure}

@@

Within R, the natural implementation for an interactive interface would
use \texttt{shiny} \citep{shinym} with \texttt{plotly} \citep{plotly}
drawing. However, the \texttt{plotly} package is designed to add to
\texttt{ggplot2} \citep{ggplot2} graphics, not to be a programmable
interactive graphics system. It lacks the capability to implement the
primary tool needed for a manual tour, which is the tracking of the
cursor position in small increments. In theory, the historical
\texttt{locator()} function of base R should provide the necessary
capability, but its behavior is not consistent between graphical
devices. The cursor position control is possible for 3D plots with
\texttt{rgl} \citep{rgl} and \texttt{plotly} but these not the arbitrary
dimension needed for tours. The level of interactivity would have been
possible with the \texttt{cranvas} system \citep{cranvas} or
\texttt{rJava} \citep{rJava} but neither has a graphics toolkit that
currently operates across platforms. What we have chosen to do is to use
the interactivity of Mathematica to experiment, and to implement the
resulting ideas as an animation in the \texttt{tourr} \citep{tourr}
package. There is hope that better interactivity will be available for R
soon, perhaps in javascript \citep[e.g.][\citet{detourr}]{langevitour},
which might allow for the experimentation to develop new techniques.

@@

The Mathematica implementation relies on the inbuilt function
\texttt{LocatorPane}. This function creates a region on the screen where
the position of the cursor is captured and then converted to input that
updates the graphics functions, enabling the manual navigation of the
tour. We then use the graphics function \texttt{ListPlot} embedded in
dynamic objects via \texttt{DynamicModule} to produce the display. User
interface objects, such as sliders, locator panes, and input fields, are
used to provide user control of slice thickness, center, scale, and
point size.

The examples presented in the application section below (and attached in
notebook format in the supplementary material) all use our primary new
function, \texttt{SliceDynamic}. This function typically accepts grouped
data in the form of a matrix where the second last column details the
name of the group and the last column details the group index (which can
be one if there is only one group). After specifying the initial slice
thickness and the slice range, the user is presented with an interactive
display in which the control objects appear on the left and the slice
visualization appears on the right. The user can change the orientation
of the slice via the locator pane, which changes the projection matrix;
a slider controls the slice thickness; and there is an input field or an
interactive star glyph, to change the slice center. The user can also
change the appearance of the plot by zooming into the center or changing
the point size with the sliders provided. It is worth noting that this
zoom works best with data that has been centered and scaled. The
projected data can be displayed to contrast it with the slice by ticking
a box. The explicit projection matrix can also be displayed via another
checkbox. This is especially useful when we wish to import a projection
that was identified as interesting in the manual exploration into a
later study, as shown in the applications below. More details about the
implementation and usage instructions are given in the Appendix.

@@

We start with a comparison of the two models and the data distribution
shifting to \(\mathbf{c}^{+}\), thus the slice is localized towards high
values of bill depth in the top row of Figure \ref{slice1p}. The star
glyph at the right indicates the slice center value, which shows that it
is higher on \texttt{bd} and at the middle value for the other
variables. This slice center guide is as proposed in \citet{condviz2},
and more details can be found in the Appendix. We can see that all three
slices (the two models and the data) contain almost no points from the
third class (Gentoo) and that the decision boundary between the two
models is very similar.

@@

A more interesting comparison is found for \(\mathbf{c}^{-}\), thus the
slice localized towards low values of bill depth, shown in the bottom
row of Figure \ref{slice1p}. (The star glyph at right indicates the
slice center value, which shows that it is lower on \texttt{bd} and at
the middle value for the other variables.) The RF model (left) predicts
all three species within this slice, with an interesting boundary for
the third class (Gentoo). On the other hand, the LDA model (middle)
predominantly predicts the third class within the slice, this appears to
be enforced through the linear structure of the model. Looking finally
at the thick slice through the data we see that there are primarily
observations from this class within the slice we can conclude that the
two models have filled in the ``empty'' space (where we do not have any
training observations) in very different ways and according to what we
might expect given the model structure.
